---
title: "Chatbot Ratings"
format: pdf
editor: visual
---

Cleaning the data below. Don't rerun this:

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(arrow) 
library(readr) 
library(ggplot2)
library(viridis)

# Cleaning data below:
# Clean_parquet <- function(data) {
#   clean_data <- data %>%
#     select(
#       -c(conversation_a, conversation_b, timestamp, evaluation_session_id, full_conversation),
#       -matches("conv_meta")
#     )
#   return(clean_data)
# }
# 
# flatten_tibble <- function(df, sep = "_") {
#   while (any(sapply(df, tibble::is_tibble))) {
#     df <- df %>%
#       unnest_wider(where(tibble::is_tibble), names_sep = sep)
#   }
#   return(df)
# }
# 
# parquet_folder <- "data"
# parquet_files <- list.files(parquet_folder, pattern = "\\.parquet$", full.names = TRUE)
# 
# all_data <- map_dfr(parquet_files, function(file) {
#   message("Processing: ", file)
#   data <- read_parquet(file)
#   data <- Clean_parquet(data)
#   data <- flatten_tibble(data)
#   return(data)
# })
# 
# write_csv(all_data, "combined_clean_flat.csv")

```

I created a grand function based on all of the work we previously did. This returns win rates, head to head match up numbers and percentages, and then the Bradley Terry model based scores for all of th e

```{r}

final_data <- read.csv("combined_clean_flat.csv")
analyze_chatbot_data <- function(data, ranking_method = "ability", 
                                 top_n = 10, plot_title = "Everything") {
  # ranking_method: 
  # "ability" for Bradley-Terry model, 
  # "win_rate" for raw win percentage
  
  # plot_title:
  # This string will go into the heat map title
  
  # Output has the following:
  # $win_rates - Outputs the win rate table
  # $abilities - Outputs the ability ranking table
  # $win_pct_matrix - Head to head percentages
  # $heatmap - Creates the heat map
  
  # ============================================================================
  # Start with Raw Win rates
  # ============================================================================
  
  processed_data <- data %>%
    mutate(model_a_orig = model_a, model_b_orig = model_b) %>%
    pivot_longer(
      cols = c(model_a, model_b),
      names_to = "model_type",
      values_to = "model"
    ) %>%
    mutate(
      model_outcome = case_when(
        winner == model_type ~ "won",
        winner %in% c("tie", "both_bad") ~ winner,
        TRUE ~ "lost"
      ),
      opponent = if_else(model_type == "model_a", model_b_orig, model_a_orig)
    ) %>%
    select(-model_a_orig, -model_b_orig, -winner, -id)
  
  processed_data <- processed_data[seq(1, nrow(processed_data), by = 2), ]
  processed_data <- filter(processed_data, !(model == opponent))
  
  model_perspective <- processed_data %>%
    rename(model_name = model) %>%
    select(model_name, model_outcome)
  
  opponent_perspective <- processed_data %>%
    rename(model_name = opponent) %>%
    mutate(model_outcome = case_when(
      model_outcome == "won" ~ "lost",
      model_outcome == "lost" ~ "won",
      model_outcome == "tie" ~ "tie",
      model_outcome == "both_bad" ~ "both_bad"
    )) %>%
    select(model_name, model_outcome)
  
  all_games <- bind_rows(model_perspective, opponent_perspective)
  
  win_rates <- all_games %>%
    group_by(model_name) %>%
    summarise(
      total_games = n(),
      wins = sum(model_outcome == "won") + sum(model_outcome == "tie"),
      losses = sum(model_outcome == "lost") + sum(model_outcome == "both_bad"),
      win_rate = (sum(model_outcome == "won") + sum(model_outcome == "tie")) / n()
    ) %>%
    arrange(desc(win_rate))
  
  # ============================================================================
  # The following section finds the head to head percentages
  # ============================================================================
  
  final_data_awin <- data %>% filter(winner == 'model_a')
  final_data_bwin <- data %>% filter(winner == 'model_b')
  
  results <- list()
  for (model in unique(final_data_awin$model_a)) {
    temp <- final_data_awin %>%
      filter(model_a == model) %>%
      count(model_b) %>%
      pivot_wider(
        names_from = model_b,
        values_from = n,
        values_fill = 0
      ) %>%
      mutate(model = model)
    results[[model]] <- temp
  }
  loss_data <- bind_rows(results)
  
  results_b <- list()
  for (model in unique(final_data_bwin$model_b)) {
    temp <- final_data_bwin %>%
      filter(model_b == model) %>%
      count(model_a) %>%
      pivot_wider(
        names_from = model_a,
        values_from = n,
        values_fill = 0
      ) %>%
      mutate(model = model)
    results_b[[model]] <- temp
  }
  loss_data_b <- bind_rows(results_b)
  
  loss_data <- loss_data %>%
    relocate(model, .before = everything())
  loss_data_b <- loss_data_b %>%
    relocate(model, .before = everything())
  
  loss_data <- loss_data %>%
    select(model, all_of(loss_data$model))
  loss_data_b <- loss_data_b %>%
    select(model, all_of(loss_data$model))
  loss_data_b <- loss_data_b %>%
    slice(match(loss_data$model, loss_data_b$model))
  
  models <- loss_data$model
  loss_data <- loss_data %>%
    select(-model)
  loss_data <- as.data.frame(sapply(loss_data, as.numeric))
  
  loss_data_b <- loss_data_b %>%
    select(-model)
  loss_data_b <- as.data.frame(sapply(loss_data_b, as.numeric))
  
  loss_data <- loss_data + loss_data_b
  
  # Covvert to percentages
  n <- nrow(loss_data)
  win_pct <- matrix(NA, n, n, dimnames = list(models, models))
  for (i in 1:n) {
    for (j in 1:n) {
      if (i != j & !is.na(loss_data[i, j]) & !is.na(loss_data[j, i])) {
        total_matches <- loss_data[i, j] + loss_data[j, i]
        if (total_matches > 0) {
          win_pct[i, j] <- loss_data[i, j] / total_matches
        } else {
          win_pct[i, j] <- NA
        }
      }
    }
  }
  
  win_pct_df <- as.data.frame(win_pct)
  win_pct_df <- cbind(models, win_pct_df)
  
  bt_data <- data %>%
    filter(winner %in% c("model_a", "model_b")) %>%
    mutate(result = ifelse(winner == "model_a", 1, 0)) %>%
    select(model_a, model_b, result)
  
  models_all <- sort(unique(c(bt_data$model_a, bt_data$model_b)))
  safe_names <- make.names(models_all)
  n_models <- length(models_all)
  n_matches <- nrow(bt_data)
  
  design_mat <- matrix(0, nrow = n_matches, ncol = n_models,
                       dimnames = list(NULL, safe_names))
  
  for (i in seq_len(n_matches)) {
    a <- bt_data$model_a[i]
    b <- bt_data$model_b[i]
    design_mat[i, make.names(a)] <- 1
    design_mat[i, make.names(b)] <- -1
  }
  
  bt_design_df <- as.data.frame(design_mat, stringsAsFactors = FALSE)
  bt_design_df$result <- as.numeric(bt_data$result)
  
  bt_manual <- glm(result ~ ., data = bt_design_df, family = binomial)
  
  coefs <- coef(bt_manual)
  coef_no_intercept <- coefs[names(coefs) != "(Intercept)"]
  
  abilities <- data.frame(
    model = models_all,
    coef_name = safe_names,
    ability = as.numeric(coef_no_intercept[safe_names])
  )
  
  abilities$ability[is.na(abilities$ability)] <- 0
  abilities$ability <- abilities$ability - mean(abilities$ability)
  abilities <- abilities %>% arrange(desc(ability))
  abilities <- select(abilities, -c(coef_name))
  
  # ============================================================================
  # Heat map creation now
  # ============================================================================
  
  if (ranking_method == "ability") {
    top_models <- abilities %>%
      head(top_n) %>%
      pull(model)
  } else {
    top_models <- win_rates %>%
      head(top_n) %>%
      pull(model_name)
  }

  row_indices <- which(rownames(win_pct) %in% top_models)
  col_indices <- which(colnames(win_pct) %in% top_models)

  top_win_pct_mat <- win_pct[row_indices, col_indices]

  heatmap_data <- as.data.frame(top_win_pct_mat) %>%
    mutate(model = rownames(.)) %>%
    pivot_longer(
      cols = -model,
      names_to = "opponent",
      values_to = "win_pct"
    )
  
  if (ranking_method == "ability") {
    heatmap_data$model <- factor(heatmap_data$model, levels = rev(top_models))
    heatmap_data$opponent <- factor(heatmap_data$opponent, levels = top_models)
    title_text <- paste0("Head-to-Head Win Rates: Top ", top_n, 
                         " Models (by Ability). Category: ", plot_title)
  } else {
    heatmap_data$model <- factor(heatmap_data$model, levels = rev(top_models))
    heatmap_data$opponent <- factor(heatmap_data$opponent, levels = top_models)
    title_text <- paste0("Head-to-Head Win Rates: Top ", top_n, 
                         " Models (by Win Rate). Category: ", plot_title)
  }
  
  p <- ggplot(heatmap_data, aes(x = opponent, y = model, fill = win_pct)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = ifelse(is.na(win_pct), "", 
                                  sprintf("%.2f", win_pct))),
              color = "white", size = 3, fontface = "bold") +
    scale_fill_viridis(option = "plasma", na.value = "grey90",
                       limits = c(0, 1),
                       labels = scales::percent) +
    labs(
      title = title_text,
      x = "Opponent",
      y = "Model (Row Wins)",
      fill = "Win %"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      legend.position = "right"
    ) +
    coord_fixed()
  
  return(list(
    win_rates = win_rates,
    abilities = abilities,
    win_pct_matrix = win_pct_df,
    heatmap = p
  ))
}


results_ability <- analyze_chatbot_data(final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Everything")

results_win_rate <- analyze_chatbot_data(final_data, 
                                        ranking_method = "win_rate", 
                                        top_n = 10, 
                                        plot_title = "Everything")

head(results_ability$abilities, 10)
```

Function to do head to heads below for a specific pair comparison

```{r}
get_model_comparison <- function(model1, model2, results_object) {
  win_matrix <- results_object$win_pct_matrix
  abilities  <- results_object$abilities

  if (!(model1 %in% win_matrix$models) | !(model2 %in% colnames(win_matrix))) {
    stop("One or both of the specified models are not found in the results object.")
  }
  
  m1_over_m2 <- win_matrix[win_matrix$models == model1, model2]
  m2_over_m1 <- win_matrix[win_matrix$models == model2, model1]
  
  ability1 <- abilities$ability[abilities$model == model1]
  ability2 <- abilities$ability[abilities$model == model2]
  
  p_m1_win <- exp(ability1) / (exp(ability1) + exp(ability2))
  p_m2_win <- 1 - p_m1_win
  
  tibble::tibble(
    Type = c("Empirical Win Rate", "Predicted (BT) Win Rate"),
    !!model1 := c(m1_over_m2, p_m1_win),
    !!model2 := c(m2_over_m1, p_m2_win)
  )
}


get_model_comparison("gemini-2.5-pro", 
                     "claude-sonnet-4-20250514-thinking-32k", 
                     results_ability)

```

Now we can simply work backwards by filtering for specific characteristics!

Sensible characteristics that I can see include:

1)  **is_code** - Whether the conversation involves code
2)  **category_tag_math_v0.1_math** - Whether the conversation involves mathematical problems
3)  **category_tag_criteria_v0.1_real_world** - Whether the conversation involves real world topics
4)  **category_tag_criteria_v0.1_problem_solving** - Whether the conversation involves strong problem solving elements
5)  **category_tag_criteria_v0.1_creativity** - Whether the conversation involves some kind of creative thinking (not writing)
6)  **category_tag_criteria_v0.1_complexity** - Whether the conversation involves complex topics and knowledge
7)  **category_tag_creative_writing_v0.1_creative_writing** - Whether the conversation involves creative writing

We're going to be filtering by ability as that makes more sense!

```{r}
temp_final_data <- filter(final_data, is_code == TRUE)
results_ability_code <- analyze_chatbot_data(temp_final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Coding")

results_ability_code$heatmap
```

```{r}
temp_final_data <- filter(final_data, category_tag_math_v0.1_math == TRUE)
results_ability_math <- analyze_chatbot_data(temp_final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Mathematics")

results_ability_math$heatmap
```

```{r}
temp_final_data <- filter(final_data, category_tag_criteria_v0.1_real_world == TRUE)
results_ability_RW <- analyze_chatbot_data(temp_final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Real World")

results_ability_RW$heatmap
```

```{r}
temp_final_data <- filter(final_data, 
                          category_tag_criteria_v0.1_problem_solving == TRUE)
results_ability_problem_solving <- analyze_chatbot_data(temp_final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Problem Solving")

results_ability_problem_solving$heatmap
```

```{r}
temp_final_data <- filter(final_data, 
                          category_tag_criteria_v0.1_creativity == TRUE)
results_ability_creativity <- analyze_chatbot_data(temp_final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Creativity")

results_ability_creativity$heatmap
```

```{r}
temp_final_data <- filter(final_data, 
                          category_tag_criteria_v0.1_complexity == TRUE)
results_ability_complex <- analyze_chatbot_data(temp_final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Complex")

results_ability_complex$heatmap
```

7)  **category_tag_creative_writing_v0.1_creative_writing** - Whether the conversation involves creative writing

```{r}
temp_final_data <- filter(final_data,
                          category_tag_creative_writing_v0.1_creative_writing == TRUE)
results_ability_creative_writing <- analyze_chatbot_data(temp_final_data, 
                                        ranking_method = "ability", 
                                        top_n = 10, 
                                        plot_title = "Creative Writing")

results_ability_creative_writing$heatmap
```

```{r}
plot_ability_scores <- function(results_object, category_name) {
  # Get top 10 models from the overall ranking, not specific
  top_10_models <- results_ability$abilities %>%
    head(10) %>%
    pull(model)
  
  plot_data <- results_object$abilities %>%
    filter(model %in% top_10_models)

  plot_data <- plot_data %>%
    mutate(model = factor(model, levels = top_10_models[top_10_models %in% plot_data$model]))

  p <- ggplot(plot_data, aes(x = model, y = ability, fill = ability)) +
    geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
    geom_text(aes(label = sprintf("%.3f", ability)), 
              vjust = -0.5, size = 3.5, fontface = "bold") +
    scale_fill_viridis(option = "plasma") +
    labs(
      title = paste0("Bradley-Terry Ability Scores: ", category_name),
      subtitle = "Top 10 Models (based on overall ranking)",
      x = "Model",
      y = "Ability Score"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 11),
      axis.text.x = element_text(angle = 60, hjust = 1, size = 14),
      axis.text.y = element_text(size = 10),
      legend.position = "none"
    )
  
  return(p)
}

plot_ability_scores(results_ability, "Everything (Overall)")
plot_ability_scores(results_ability_code, "Coding")
plot_ability_scores(results_ability_math, "Mathematics")
plot_ability_scores(results_ability_RW, "Real World")
plot_ability_scores(results_ability_problem_solving, "Problem Solving")
plot_ability_scores(results_ability_creativity, "Creativity")
plot_ability_scores(results_ability_complex, "Complex")
plot_ability_scores(results_ability_creative_writing, "Creative Writing")
```
